Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2020-11-29 numerics_paper/version2/numerical_chj.tex
Put back missing subscripts on entropy potential.

\subsection{The circular hydraulic jump in two dimensions}\label{sec:2D_chj}
Let us finally consider the numerical experiments for the CHJ in two-dimensions.
The domain is given by the annulus $\Omega=\{(x,y)~|~ 0.1\leq\sqrt{x^2+y^2}\leq 1\}$.
The computational mesh is a logically quadrilateral mesh, see Figure \ref{fig:mesh_chj}.
In all the experiments in this section, the mesh consists of $10^6$ cells. 
Let
\begin{align*}
  \partial\Omega_I:=\{(x,y)\in\Omega ~|~ \sqrt{x^2+y^2}=0.1\}, \qquad \mbox{ and } \qquad
  \partial\Omega_o:=\{(x,y)\in\Omega ~|~ \sqrt{x^2+y^2}=1\},
\end{align*}
denote the inflow and outflow boundaries, respectively. 
The boundary conditions are 
\begin{subequations}\label{bcs_chj}
  \begin{align}
  h(x,y,t)=h_I, \quad
  u(x,y,t)=|\bfu_I| \left(\frac{x}{r_I}\right), \quad
  v(x,y,t)=|\bfu_I| \left(\frac{y}{r_I}\right), \qquad \forall (x,y)\in\partial\Omega_I, \\
  %
  h(x,y,t)=h_o, \quad 
  u(x,y,t)= \frac{\beta }{r_o h}\left(\frac{x}{r_o}\right), \quad
  v(x,y,t)= \frac{\beta }{r_o h}\left(\frac{y}{r_o}\right), \qquad \forall (x,y)\in\partial\Omega_o, \label{bcs_chj_outflow}
  \end{align}
\end{subequations}
where $h_I=0.3$, $r_I=0.1$, $r_o=1$ and $\beta=r_Ih_I|\bfu_I|$.
We choose $|\bfu_I|$ and $h_o$ to generate a radially symmetric shock at $r_s:=\sqrt{x^2+y^2}=0.3$, as described in \S \ref{sec:steady_chj}.
By selecting the boundary conditions $|\bfu_I|$ and $h_o$ we can move through different regimes.
In particular, we consider two cases.
For the first case, the Froude number at the inflow boundary is $\Fr\approx 1.3693$.
In this case, the CHJ appears to be nearly stable.
However, using Roe's solver leads to the formation of carbuncle instabilities.
%
The second regime is a more aggresive regime with a Froude number of $\Fr\approx 27.3861$ at the inflow boundary. 
In this regime, using Roe's solver also leads to carbuncle instabilities. In addition, 
due to the large Froude number, random perturbations might trigger physical instabilities. 
If the random perturbations model or mimic real disturbances in a physical experiment, it is then
desirable to properly simulate the unstable behavior of the flow without developing carbuncle instabilities. 
Therefore, the numerical method must introduce a carbuncle fix without dissipating important features of the
solution that could lead to physical instabilities. 
For this reason, we believe this problem is more challenging than the standard bow shock problem.

@@

\subsubsection{Regime I}
In this case, the boundary conditions are given by \eqref{bcs_chj} with
\begin{align}\label{bcs_chj_r1}
  |\bfu_I|=0.75, \qquad h_o=0.37387387318873766. 
\end{align}
The initial condition is a symmetric CHJ located at $\sqrt{x^2+y^2}=0.3$, generated as described in \S \ref{sec:steady_chj}.
In Figure \ref{fig:chj_r1}, we show the magnitude of the momentum at different times using
method \eqref{second-order_via_fluct} with Roe's Riemann solver.
The solution clearly develops carbuncle instabilities.
We could fix these instabilities using Rusanov's, Kemm's or the blended Riemann solvers. 
In Figure \ref{fig:chj_r1_diff}, we show the corresponding solutions at $t=3$.
For this test case, all solvers give qualitatively similar solutions;
however, we expect Rusanov's solver to be less accurate.

\begin{figure}[!h]
  \centering 
  \includegraphics[scale=0.28]{figures/chj_r1_roe_p1.png} \quad
  \includegraphics[scale=0.28]{figures/chj_r1_roe_p2.png} \quad
  \includegraphics[scale=0.28]{figures/chj_r1_roe_p3.png} \quad
  \caption{Magnitude of the momentum for a CHJ with boundary conditions given by
    \eqref{bcs_chj} and \eqref{bcs_chj_r1}. 
    We consider method \eqref{second-order_via_fluct} with Roe's solver.    
    The color scale in the plots go from (blue) $0.0746$ to (red) $0.3738$,
    which corresponds to the minimum and maximum values of the momentum at $t=0$.
    \label{fig:chj_r1}}
\end{figure}

\begin{figure}[!h]
  \centering 
  \includegraphics[scale=0.28]{figures/chj_r1_llf_p3.png} \quad
  \includegraphics[scale=0.5]{figures/cross.png} \quad
  \includegraphics[scale=0.28]{figures/chj_r1_blended_p3.png} \quad
  \caption{Magnitude of the momentum for a CHJ with boundary conditions given by
    \eqref{bcs_chj} and \eqref{bcs_chj_r1}. 
    We consider method \eqref{second-order_via_fluct} with (left) Rusanov's,
    (middle) Kemm's and (right) the blended solvers. 
    The color scale in the plots go from (blue) $0.0746$ to (red) $0.3738$,
    which corresponds to the minimum and maximum values of the momentum at $t=0$.
    \label{fig:chj_r1_diff}}
\end{figure}

\subsubsection{Regime II}
We now consider a more aggresive regime. The boundary conditions are given by \eqref{bcs_chj} with 
\begin{align}\label{bcs_chj_r2}
  |\bfu_I|=15, \qquad h_o=6.6845019298155357.
\end{align}
Similar than before, we follow \S \ref{sec:steady_chj} to generate an initial condition with a symmetric CHJ located at
$\sqrt{x^2+y^2}=0.3$. In Figure XX, we show the magnitude of the momentum at different times using method
\eqref{second-order_via_fluct} with Roe's Riemann solver. Just like before, the solution develops
carbuncle instabilities. Let us now use Rusanov's, Kemm's and the blended solvers to fix the
carbuncle instabilities. In Figure XX, we show the corresponding solutions at $t=XX$.
{\color{red} TODO: finish this part depending on the results with Kemm's solver.}
%
Opposed to what we observed in the first test case, in this situation the Froude number is
high enough that numerical perturbations start to trigger a response around the CHJ.
For the resolution that we consider, this behavior is dissipated by the extra numerical viscosity introduced by
Rusanov's solver. In Figure XX, we show the solution at a later time considering the same solvers as before.
Clearly, using the Rusanov's solver dissipated the physical response to these numerical perturbations.

\begin{figure}[!h]
  \centering 
  \includegraphics[scale=0.28]{figures/chj_r2_roe_p1.png} \quad
  \includegraphics[scale=0.28]{figures/chj_r2_roe_p2.png} \quad
  \includegraphics[scale=0.28]{figures/chj_r2_roe_p3.png}
  \caption{Magnitude of the momentum for a CHJ with boundary conditions given by
    \eqref{bcs_chj} and \eqref{bcs_chj_r2}. 
    We consider method \eqref{second-order_via_fluct} with Roe's solver.
    The color scale in the plots go from (blue) $0.4502$ to (red) $4.4798$,
    which corresponds to the minimum and maximum values of the momentum at $t=0$.
    \label{fig:chj_r2}}
\end{figure}

\begin{figure}[!h]
  \centering 
  \includegraphics[scale=0.28]{figures/chj_r2_llf_t0p1.png} \quad
  \includegraphics[scale=0.5]{figures/cross.png} \quad
  \includegraphics[scale=0.28]{figures/chj_r2_blended_t0p1.png} 
  \caption{Magnitude of the momentum for a CHJ with boundary conditions given by
    \eqref{bcs_chj} and \eqref{bcs_chj_r2}. 
    We consider method \eqref{second-order_via_fluct} with (left) Rusanov's,
    (middle) Kemm's and (right) the blended solvers. 
    The color scale in the plots go from (blue) $0.4502$ to (red) $4.4798$,
    which corresponds to the minimum and maximum values of the momentum at $t=0$.
    \label{fig:chj_r2}}
\end{figure}

\begin{figure}[!h]
  \centering 
  \includegraphics[scale=0.5]{figures/cross.png} \quad
  \includegraphics[scale=0.5]{figures/cross.png} \quad
  \includegraphics[scale=0.28]{figures/chj_r2_blended_t5p0.png} 
  \caption{Magnitude of the momentum for a CHJ with boundary conditions given by
    \eqref{bcs_chj} and \eqref{bcs_chj_r2}. 
    We consider method \eqref{second-order_via_fluct} with (left) Rusanov's,
    (middle) Kemm's and (right) the blended solvers. 
    The color scale in the plots go from (blue) $0.4502$ to (red) $4.4798$,
    which corresponds to the minimum and maximum values of the momentum at $t=0$.
    \label{fig:chj_r2}}
\end{figure}

\subsubsection{Regime II with random perturbations}
Finally, we introduce a physical random perturbation to the inflow boundary condition.
Let $\epsilon(x,y,t)\in[-0.01,0.01]$ be a random number for a given location $(x,y)$,
at a given time $t$. 
The inflow boundary condition is now given by
\begin{subequations}
  \begin{align}
    h(x,y,t)=\tilde h_I, \quad
    u(x,y,t)=|\tilde \bfu_I| \left(\frac{x}{r_I}\right), \quad
    v(x,y,t)=|\tilde \bfu_I| \left(\frac{y}{r_I}\right), \qquad \forall (x,y)\in\partial\Omega_I,
  \end{align}
\end{subequations}
with $\tilde h_I = \frac{h_I}{1+\epsilon(x,y)}$ and $|\tilde\bfu_I|=|\bfu_I|(1+\epsilon(x,y))$.
Here, just as before, $h_I=0.3$ and $|\bfu_I|=15$. 
Note that $\beta=r_I\tilde h|\tilde\bfu_I|=r_Ih_I\bfu_I$.
The outflow boundary condition is given by
\eqref{bcs_chj_outflow} with $h_o=6.6845019298155357$.

\begin{figure}[!h]
  \centering
  \begin{tabular}{cccc}
    \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t1p0.png} &
    \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t2p0.png} &
    \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t4p0.png} &
    \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t6p0.png}
  \end{tabular}
  \caption{XX}
\end{figure}

>>>>

\subsection{The circular hydraulic jump in two dimensions}\label{sec:2D_chj}
Let us finally consider the numerical experiments for the CHJ in two-dimensions.
The domain is given by the annulus $\Omega=\{(x,y)~|~ 0.1\leq\sqrt{x^2+y^2}\leq 1\}$.
The computational mesh is a logically quadrilateral mesh, see Figure \ref{fig:mesh_chj}.
For all the experiments in this section, the mesh consists of $10^6$ cells. 
Let
\begin{align*}
  \partial\Omega_I:=\{(x,y)\in\Omega ~|~ \sqrt{x^2+y^2}=0.1\}, \qquad \mbox{ and } \qquad
  \partial\Omega_o:=\{(x,y)\in\Omega ~|~ \sqrt{x^2+y^2}=1\},
\end{align*}
denote the inflow and outflow boundaries, respectively. 
The boundary conditions are 
\begin{subequations}\label{bcs_chj}
  \begin{align}
  h(x,y,t)=h_I, \quad
  u(x,y,t)=|\bfu_I| \left(\frac{x}{r_I}\right), \quad
  v(x,y,t)=|\bfu_I| \left(\frac{y}{r_I}\right), \qquad \forall (x,y)\in\partial\Omega_I, \\
  %
  h(x,y,t)=h_o, \quad 
  u(x,y,t)= \frac{\beta }{r_o h}\left(\frac{x}{r_o}\right), \quad
  v(x,y,t)= \frac{\beta }{r_o h}\left(\frac{y}{r_o}\right), \qquad \forall (x,y)\in\partial\Omega_o, \label{bcs_chj_outflow}
  \end{align}
\end{subequations}
where $h_I=0.3$, $r_I=0.1$, $r_o=1$ and $\beta=r_Ih_I|\bfu_I|$.
We choose $|\bfu_I|$ and $h_o$ to generate a radially symmetric shock at $r_s:=\sqrt{x^2+y^2}=0.3$, as described in \S \ref{sec:steady_chj}.
By selecting the boundary conditions $|\bfu_I|$ and $h_o$ we can move through different regimes.
In particular, we consider two cases.
For the first case, the Froude number at the inflow boundary is $\Fr\approx 1.3693$.
In this case, the CHJ appears to be nearly stable.
However, using Roe's solver leads to the formation of carbuncle instabilities.
%
The second regime is a more aggresive regime with a Froude number of $\Fr\approx 27.3861$ at the inflow boundary. 
In this regime, using Roe's solver also leads to carbuncle instabilities. In addition, 
due to the large Froude number, random perturbations might trigger physical instabilities. 
If the random perturbations model or mimic real disturbances in a physical experiment, it is then
desirable to properly simulate the unstable behavior of the flow without developing carbuncle instabilities. 
Therefore, the numerical method must introduce a carbuncle fix without dissipating important features of the
solution that could lead to physical instabilities. 
For this reason, we believe this problem is more challenging than the standard bow shock problem.

For most of the experiments we show a Schlieren plot for the water height.
That is, we plot $||\nabla h||_{\ell^2}$ with a white-to-black logarithm based colormap.
For some experiments we also plot the magnitude of the momentum $|h\bfu|=h\sqrt{u^2+v^2}$.

@@

\subsubsection{Regime I}
In this case, the boundary conditions are given by \eqref{bcs_chj} with
\begin{align}\label{bcs_chj_r1}
  |\bfu_I|=0.75, \qquad h_o=0.37387387318873766. 
\end{align}
The initial condition is a symmetric CHJ located at $\sqrt{x^2+y^2}=0.3$, generated as described in \S \ref{sec:steady_chj}.
In Figure \ref{fig:chj_r1_roe}, we show the solution at different times using
method \eqref{second-order_via_fluct} with Roe's Riemann solver.
The solution clearly develops carbuncle instabilities.
We could fix these instabilities using Rusanov's, Kemm's or the blended Riemann solvers. 
In Figure \ref{fig:chj_r1_diff}, we show the corresponding solutions at $t=3$.
For this test case, all carbuncle free solvers give qualitatively similar solutions;
however, we expect Rusanov's solver to be less accurate.

\begin{figure}[!h]
  \centering
  \subfloat[Solution at (left) $t=1$, (middle) $t=1.5$ and (right) $t=3$ using Roe's Riemann solver.\label{fig:chj_r1_roe}]{
    \begin{tabular}{ccc} 
      \includegraphics[scale=0.3]{figures/chj_r1_roe_p1.png} &
      \includegraphics[scale=0.3]{figures/chj_r1_roe_p2.png} &
      \includegraphics[scale=0.3]{figures/chj_r1_roe_p3.png}
    \end{tabular}
  }

  \subfloat[Solution at $t=3$ using (left) Rusanov's, (middle) Kemm's and (right) the blended solvers.\label{fig:chj_r1_diff}]{
    \begin{tabular}{ccc}
      \includegraphics[scale=0.3]{figures/chj_r1_llf_p3.png} &
      \includegraphics[scale=0.5]{figures/cross.png} &
      \includegraphics[scale=0.3]{figures/chj_r1_blended_p3.png}
    \end{tabular}
    }
    \caption{CHJ with boundary conditions given by \eqref{bcs_chj} and \eqref{bcs_chj_r1}.    
      We consider method \eqref{second-order_via_fluct} with different Riemann solvers.
      In all cases we show the Schlieren plot for the water height $h$.
    %The color scale in the plots go from (blue) $0.0746$ to (red) $0.3738$,
    %which corresponds to the minimum and maximum values of the momentum at $t=0$.
    \label{fig:chj_r1}}
\end{figure}

\subsubsection{Regime II}
We now consider a more aggresive regime. The boundary conditions are given by \eqref{bcs_chj} with 
\begin{align}\label{bcs_chj_r2}
  |\bfu_I|=15, \qquad h_o=6.6845019298155357.
\end{align}
Similar than before, we follow \S \ref{sec:steady_chj} to generate an initial condition with a symmetric CHJ located at
$\sqrt{x^2+y^2}=0.3$. In Figure \ref{fig:chj_r2_roe}, we show the solution at different times using method
\eqref{second-order_via_fluct} with Roe's Riemann solver. Just like before, the solution develops
carbuncle instabilities. Let us now use Rusanov's, Kemm's and the blended solvers to fix the
carbuncle instabilities. In Figure \ref{fig:chj_r2_diff}, we show the corresponding solutions at $t=0.1$.
{\color{red} TODO: finish this part depending on the results with Kemm's solver.}
%
Opposed to what we observed in the first test case, in this situation the Froude number is
high enough that numerical perturbations start to trigger a response around the CHJ
(see the middle and right panels of Figure \ref{fig:chj_r2_diff}).
For the resolution that we consider, this behavior is dissipated by the extra numerical viscosity introduced by
Rusanov's solver (see the left panel of Figure \ref{fig:chj_r2_diff}).
In Figure \ref{fig:chj_r2_later}, we show the solution at a later time considering the same solvers as before.
This time, we plot not only the Schlieren plot for the water height, but also the magnitude of the momentum. 
Clearly, using Rusanov's solver dissipates the physical response to these numerical perturbations.

\begin{figure}[!h]
  \centering
  \subfloat[Solution at (left) $t=0.09$, (middle) $t=0.1$ and (right) $t=0.11$ using Roe's Riemann solver.\label{fig:chj_r2_roe}]{
    \begin{tabular}{ccc}
      \includegraphics[scale=0.3]{figures/chj_r2_roe_p1.png} &
      \quad\includegraphics[scale=0.3]{figures/chj_r2_roe_p2.png} \quad &
      \includegraphics[scale=0.3]{figures/chj_r2_roe_p3.png}
    \end{tabular}
  }

  \vspace{10pt}
  \subfloat[Solution at $t=0.1$ using (left) Rusanov's, (middle) Kemm's and (right) the blended solvers.\label{fig:chj_r2_diff}]{
    \begin{tabular}{ccc}
      \includegraphics[scale=0.3]{figures/chj_r2_llf_t0p1.png} &
      \quad \includegraphics[scale=0.5]{figures/cross.png} \quad &
      \includegraphics[scale=0.3]{figures/chj_r2_blended_t0p1.png}
    \end{tabular}
    }
  \caption{CHJ with boundary conditions given by \eqref{bcs_chj} and \eqref{bcs_chj_r2}. 
    We consider method \eqref{second-order_via_fluct} with different Riemann solvers.
    In all cases we show the Schlieren plot for the water height $h$.
    %The color scale in the plots go from (blue) $0.45$ to (red) $4.5$,
    %which corresponds to the minimum and maximum values of the momentum at $t=0$.
  }
\end{figure}

\begin{figure}[!h]
  \centering
  \subfloat[$t=5$]{
    \includegraphics[scale=0.3]{figures/chj_r2_rusanov_t5p0.png} \quad
    \includegraphics[scale=0.5]{figures/cross.png} \quad
    \includegraphics[scale=0.3]{figures/chj_r2_blended_t5p0.png}
  }
  \caption{CHJ at $t=5$ with boundary conditions given by \eqref{bcs_chj} and \eqref{bcs_chj_r2}. 
    We consider method \eqref{second-order_via_fluct} with (left) Rusanov's,
    (middle) Kemm's and (right) the blended solvers. 
    In all cases we superimpose the Schlieren plot for the water height $h$
    an the magnitude of the momentum with a uniform (white-to-blue) colormap. 
    The scale for colormap is (white) $0.45$ to (blue) $4.5$, 
    which corresponds to the minimum and maximum values at $t=0$.
    \label{fig:chj_r2_later}}
\end{figure}

\subsubsection{Regime II with random perturbations}
Finally, we introduce a physical random perturbation to the inflow boundary condition.
Let $\epsilon(x,y,t)\in[-0.01,0.01]$ be a random number for a given location $(x,y)$,
at a given time $t$. 
The inflow boundary condition is now given by
\begin{subequations}
  \begin{align}
    h(x,y,t)=\tilde h_I, \quad
    u(x,y,t)=|\tilde \bfu_I| \left(\frac{x}{r_I}\right), \quad
    v(x,y,t)=|\tilde \bfu_I| \left(\frac{y}{r_I}\right), \qquad \forall (x,y)\in\partial\Omega_I,
  \end{align}
\end{subequations}
with $\tilde h_I = \frac{h_I}{1+\epsilon(x,y,t)}$ and $|\tilde\bfu_I|=|\bfu_I|(1+\epsilon(x,y,t))$.
Here, just as before, $h_I=0.3$ and $|\bfu_I|=15$. 
Note that $\beta=r_I\tilde h|\tilde\bfu_I|=r_Ih_I\bfu_I$.
The outflow boundary condition is given by
\eqref{bcs_chj_outflow} with $h_o=6.6845019298155357$.
In this case, due to the large Froude number, the random perturbations in the inflow
might trigger physical instabilities, which we want to preserve. At the same time, we
do not want the solution to develop carbuncle instabilities.
In figures \ref{fig:chj_r2wp_rusanov} and \ref{fig:chj_r2wp_blended}, we show the solution
at different times considering method \eqref{second-order_via_fluct} with Rusanov's and
the blended Riemann solvers, respectively.

\begin{figure}[!h]
  \centering
  \subfloat[Rusanov's Riemann solver\label{fig:chj_r2wp_rusanov}]{
    \begin{tabular}{cccc}
      $t=2$ & $t=3$ & $t=4$ & $t=5$ \\
      \includegraphics[scale=0.24]{figures/chj_r2wp_rusanov_t2p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_rusanov_t3p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_rusanov_t4p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_rusanov_t5p0.png}
    \end{tabular}
  }

  \subfloat[Blended Riemann solver\label{fig:chj_r2wp_blended}]{
    \begin{tabular}{cccc}
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t2p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t3p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t4p0.png} &
      \includegraphics[scale=0.24]{figures/chj_r2wp_blended_t5p0.png}
    \end{tabular}
  }
  \caption{CHJ at different times with a random perturbation at the inflow boundary.
    We consider method \eqref{second-order_via_fluct} with (a) Roe's and (b) the blended
    Riemann solvers. In all cases we show the Schlieren plot for the water height $h$.}
\end{figure}
